\sysfull

### 2018-04-10 paper/intro.tex
change of cite format

Advances in processor efficiency along with the development of energy-harvesting systems has created a new category of devices that require neither a battery nor a tethered power supply~\cite{prasad_comst_2014,lucia_snapl_2017,soyata_csm_2016}. These devices operate using ambient energy, such as radio frequency transmissions~\cite{rf_powered_computing_gollakota_2014}, light~\cite{margolies_infocom_2016,margolies_tosn_2016}, and vibration~\cite{gorlatova_sigmetrics_2014}. Incorporating compute, storage, sensing, and communication hardware~\cite{wisp5,moo}, such devices are a promising technology for use in the Internet of Things~\cite{ku_cst_2016}, in-body~\cite{nadeau_naturebio_2017} and on-body~\cite{bandodkar_electroanalysis_2015} medical systems, and energy-harvesting nano-satellites~\cite{kicksat}.

Energy-harvesting devices create unique challenges because they operate {\em intermittently} when energy is available~\cite{hicks_isca_2017,lucia_snapl_2017}. An
energy-harvesting device buffers energy in a small storage, e.g. a capacitor~\cite{gorlatova_tmc_2013},~\cite{gunduz_commag_2014} and when a threshold amount of energy accumulates, begins operating. Harvestable energy sources are low-power compared to a platform's operating level. A device operates briefly and when buffered energy is depleted, shuts down and recharges to operate again later. As an example, recharge time may be tens of seconds in radio frequency powered medical device~\cite[Fig. 3c]{nadeau_naturebio_2017}. Moreover, charge and discharge times vary by device (due to different capacitor sizes) and some may fail $\approx$10 to $\approx$100 times per second~\cite{tan_infocom_2016},~\cite{mementos},~\cite{nvp}.

%\begin{figure}
%	\begin{subfigure}[t]{.35\linewidth}
%		\centering\large Figure
%		\caption{\centering WISP~\cite{wisp} against an RFID antenna}\label{fig:1a}
%	\end{subfigure}%
%	\begin{subfigure}[t]{.65\linewidth}
%		\centering\large Figure
%		\caption{\centering Program X~\cite{hicks_mibench2_2016} (split into X and Y tasks~\cite{chain}) at two distances}\label{fig:1b}
%    \end{subfigure}
%	\caption{Impact of an inadequate task sizing on the computation speed of a moving transiently-powered device.}\label{fig:1}
%\end{figure}

@@

\textbf{Data Consistency in Intermittent Computing.} Software in an energy-harvesting system operates in the {\em intermittent execution model}~\cite{dino,lucia_snapl_2017}, with execution interrupted by failure periods. When power fails, a device loses volatile state, i.e., registers, stack, SRAM, and retains its non-volatile state, i.e., FRAM, Flash. While capturing periodic checkpoints~\cite{mementos,quickrecall} and sleep scheduling~\cite{dewdrop,hibernus,hibernusplusplus} help preserve execution progress, failures can leave volatile and non-volatile state inconsistent, leading to unrecoverable failures~\cite{dino,edb}.

There are two main approaches for dealing with data inconsistency for intermittently-powered devices: (i) \emph{programming and execution models}~\cite{dino,ratchet,chain,alpaca} and (ii) \emph{architectures}~\cite{hicks_isca_2017,idetic,nvp,tictpl}. New architectures require hardware changes and are inapplicable to today's systems~\cite{hicks_isca_2017,nvp}, therefore new programming models and compilers are more suitable (as of now). However, they have their own limitations that need to be addressed---the core one being the rigidity of the \emph{static} decomposition into tasks.

@@

\textbf{Task Decomposition of Intermittent Programs.} Recently proposed {\em
task-based} programming and execution models~\cite{alpaca,chain} advocate for
static decomposition of a program (by a programmer) into a collection of tasks.
Tasks can include arbitrary computation and, upon completion, are guaranteed to
have executed {\em atomically}, despite arbitrarily-timed power failures.
The programmer explicitly expresses task-to-task control flow. The execution
in Figure~\ref{fig:coalesce} (top) illustrates how the decomposition into
tasks affects the time the application takes to complete. Each task transition introduces the overhead of tracking and atomically committing the modifications to non-volatile memory, to maintain consistency of program state~\cite{chain,alpaca}. Naively minimizing the number of task transitions, (i.e. the task count), \emph{at compile time} risks creating a task that requires more energy than the device can buffer in its energy storage. Such a large task would fail to complete unless the environment provides sufficient incoming energy to supplement the energy in the capacitor. To eliminate this risk, the programmer (or the compiler) must decompose the program conservatively into many small tasks, at the cost of \emph{higher task transition overhead}.

@@

%A key challenge is that the length of a software
%task's execution is \emph{limited by the fixed total amount of energy} that a
%device can buffer in hardware. A task's code is static, but the duration of its
%execution may be input-dependent and is difficult to predict. To illustrate
%this, we refer to Figure~\ref{fig:1} showing the execution time of two
%applications (the same application X, but divided into X and Y tasks as
%in~\cite{chain}) running on RFID antenna-powered Computational
%RFID~\cite{wisp,rf_powered_computing_gollakota_2014} at two device-to-antenna
%distances (near---stable energy supply/far---high energy intermittency). At
%short distance: execution will takes too long caused by runtime cost of
%marshalling excessively small tasks. At far distance: program \emph{might never
%execute} if task execution consumes more energy than the system can buffer.
%This calls for at-runtime adaptive task division of any transiently-powered
%application---namely \emph{task virtualization}.

@@

Unfortunately, naively merging two atomic tasks might produce a non-atomic task, that may leave non-volatile memory inconsistent after a partial execution. A merge breaks atomicity when atomic merged tasks form a \emph{write-after-read (WAR)} dependency---for instance if two tasks \texttt{\{x=y+1\}; \{y=x;\}} are merged and if the power failure occurs after \texttt{x=y}, the value \texttt{x} will be increased twice when the merged task is restarted, that leads to an inconsistency.
%
%variables in non-volatile memory at \emph{run-time} that can
%break the atomicity of the \emph{virtual task}. Consider the
%example depicted in Fig.~\ref{fig:virtualization}: Tasks 1,
%2 and 3 being executed consecutively. All tasks in this
%example are atomic since they do not have WAR dependency on
%the persistent variables they are accessing, e.g. \emph{x}
%is only read and \emph{y} is only written within Task 1.
%Now, suppose three tasks have been virtualized into a single
%one at run-time, namely Task 4: since \emph{x} is now first
%\emph{read} and then \emph{written}, a WAR dependency on
%\emph{x} is introduced dynamically at
%run-time---unfortunately Task 4 is \emph{no more atomic}
%since its re-execution will not always produce the same
%results.
%
Although WAR dependencies introduced by the \emph{static} task division can be eliminated at compile time~\cite{alpaca}, those introduced \emph{dynamically} at
run-time cannot. To maintain memory consistency in merged tasks, we propose a
mechanism for \emph{memory virtualization} that buffers the updates to non-volatile variables in volatile memory before committing them at the dynamic task boundary.
%
%Therefore, we require a \emph{new execution
%model} that keeps each virtual task atomic by committing the
%modified persistent variables at the boundary defined
%dynamically at run-time---keeping the non-volatile memory
%unmodified upon a power-interrupt and preserving its
%consistency.

@@

%Existing programming and execution models may require more accesses to non-volatile memory (for multi-versioning)~\cite{dino,chain} or may preclude the use of volatile memory~\cite{ratchet}. Effective use of both volatile and non-volatile memory is important for \emph{efficiency}, because non-volatile memory has higher access energy and latency than SRAM~\cite{nvp}, and \emph{generality}, because devices often have much more FRAM than SRAM (e.g., 64 times more in~\cite{wolverine}). How to connect this trade-off with task virtualization remains an open question.

@@

We evaluated \sys end-to-end on a real energy-harvesting device~\cite{wisp} running existing benchmarks ~\cite{chain} and new programs. A comparison to a state-of-the-art task-based system~\cite{chain} showed that \sys recovers
performance on conservative task decompositions missed by systems with static decomposition.
%
%and can flexibly target a variety of platforms without rewriting code.
%
%\todo{Provide concrete numeric results supporting this claim}{Amjad/Przemek}

%In this work we also address the problem of task construction burden imposed on the programmer. That is, we introduce \sys's compiler that transforms arbitrary C code into a graph of tasks that reflects the original program's control- and data-flow constraints. The compiler identifies data shared by multiple of the newly created tasks and instruments reads and writes of those data so that \sys's memory virtualization mechanism keeps those data consistent. While \sys's complier is still \emph{limited in scope}, e.g. is unable to handle instruction jumps, normalized loops and recursions, it already eliminates the labour-intensive process of writing tasks~\cite{chain,alpaca}, i.e., hours of programmer time.

>>>>

Advances in processor efficiency along with the development of energy-harvesting systems has created a new category of devices that require neither a battery nor a tethered power supply~\citep{prasad_comst_2014,lucia_snapl_2017,soyata_csm_2016}. These devices operate using ambient energy, such as radio frequency transmissions~\citep{rf_powered_computing_gollakota_2014}, light~\citep{margolies_infocom_2016,margolies_tosn_2016}, and vibration~\citep{gorlatova_sigmetrics_2014}. Incorporating compute, storage, sensing, and communication hardware~\citep{wisp5,moo}, such devices are a promising technology for use in the Internet of Things~\citep{ku_cst_2016}, in-body~\citep{nadeau_naturebio_2017} and on-body~\citep{bandodkar_electroanalysis_2015} medical systems, and energy-harvesting nano-satellites~\citep{kicksat}.

Energy-harvesting devices create unique challenges because they operate {\em intermittently} when energy is available~\citep{hicks_isca_2017,lucia_snapl_2017}. An
energy-harvesting device buffers energy in a small storage, e.g. a capacitor~\citep{gorlatova_tmc_2013},~\citep{gunduz_commag_2014} and when a threshold amount of energy accumulates, begins operating. Harvestable energy sources are low-power compared to a platform's operating level. A device operates briefly and when buffered energy is depleted, shuts down and recharges to operate again later. As an example, recharge time may be tens of seconds in radio frequency powered medical device~\cite[Fig. 3c]{nadeau_naturebio_2017}. Moreover, charge and discharge times vary by device (due to different capacitor sizes) and some may fail $\approx$10 to $\approx$100 times per second~\citep{tan_infocom_2016},~\citep{mementos},~\citep{nvp}.

%\begin{figure}
%	\begin{subfigure}[t]{.35\linewidth}
%		\centering\large Figure
%		\caption{\centering WISP~\citep{wisp} against an RFID antenna}\label{fig:1a}
%	\end{subfigure}%
%	\begin{subfigure}[t]{.65\linewidth}
%		\centering\large Figure
%		\caption{\centering Program X~\citep{hicks_mibench2_2016} (split into X and Y tasks~\citep{chain}) at two distances}\label{fig:1b}
%    \end{subfigure}
%	\caption{Impact of an inadequate task sizing on the computation speed of a moving transiently-powered device.}\label{fig:1}
%\end{figure}

@@

\textbf{Data Consistency in Intermittent Computing.} Software in an energy-harvesting system operates in the {\em intermittent execution model}~\citep{dino,lucia_snapl_2017}, with execution interrupted by failure periods. When power fails, a device loses volatile state, i.e., registers, stack, SRAM, and retains its non-volatile state, i.e., FRAM, Flash. While capturing periodic checkpoints~\citep{mementos,quickrecall} and sleep scheduling~\citep{dewdrop,hibernus,hibernusplusplus} help preserve execution progress, failures can leave volatile and non-volatile state inconsistent, leading to unrecoverable failures~\citep{dino,edb}.

There are two main approaches for dealing with data inconsistency for intermittently-powered devices: (i) \emph{programming and execution models}~\citep{dino,ratchet,chain,alpaca} and (ii) \emph{architectures}~\citep{hicks_isca_2017,idetic,nvp,tictpl}. New architectures require hardware changes and are inapplicable to today's systems~\citep{hicks_isca_2017,nvp}, therefore new programming models and compilers are more suitable (as of now). However, they have their own limitations that need to be addressed---the core one being the rigidity of the \emph{static} decomposition into tasks.

@@

\textbf{Task Decomposition of Intermittent Programs.} Recently proposed {\em
task-based} programming and execution models~\citep{alpaca,chain} advocate for
static decomposition of a program (by a programmer) into a collection of tasks.
Tasks can include arbitrary computation and, upon completion, are guaranteed to
have executed {\em atomically}, despite arbitrarily-timed power failures.
The programmer explicitly expresses task-to-task control flow. The execution
in Figure~\ref{fig:coalesce} (top) illustrates how the decomposition into
tasks affects the time the application takes to complete. Each task transition introduces the overhead of tracking and atomically committing the modifications to non-volatile memory, to maintain consistency of program state~\citep{chain,alpaca}. Naively minimizing the number of task transitions, (i.e. the task count), \emph{at compile time} risks creating a task that requires more energy than the device can buffer in its energy storage. Such a large task would fail to complete unless the environment provides sufficient incoming energy to supplement the energy in the capacitor. To eliminate this risk, the programmer (or the compiler) must decompose the program conservatively into many small tasks, at the cost of \emph{higher task transition overhead}.

@@

%A key challenge is that the length of a software
%task's execution is \emph{limited by the fixed total amount of energy} that a
%device can buffer in hardware. A task's code is static, but the duration of its
%execution may be input-dependent and is difficult to predict. To illustrate
%this, we refer to Figure~\ref{fig:1} showing the execution time of two
%applications (the same application X, but divided into X and Y tasks as
%in~\citep{chain}) running on RFID antenna-powered Computational
%RFID~\citep{wisp,rf_powered_computing_gollakota_2014} at two device-to-antenna
%distances (near---stable energy supply/far---high energy intermittency). At
%short distance: execution will takes too long caused by runtime cost of
%marshalling excessively small tasks. At far distance: program \emph{might never
%execute} if task execution consumes more energy than the system can buffer.
%This calls for at-runtime adaptive task division of any transiently-powered
%application---namely \emph{task virtualization}.

@@

Unfortunately, naively merging two atomic tasks might produce a non-atomic task, that may leave non-volatile memory inconsistent after a partial execution. A merge breaks atomicity when atomic merged tasks form a \emph{write-after-read (WAR)} dependency---for instance if two tasks \texttt{\{x=y+1\}; \{y=x;\}} are merged and if the power failure occurs after \texttt{x=y}, the value \texttt{x} will be increased twice when the merged task is restarted, that leads to an inconsistency.
%
%variables in non-volatile memory at \emph{run-time} that can
%break the atomicity of the \emph{virtual task}. Consider the
%example depicted in Fig.~\ref{fig:virtualization}: Tasks 1,
%2 and 3 being executed consecutively. All tasks in this
%example are atomic since they do not have WAR dependency on
%the persistent variables they are accessing, e.g. \emph{x}
%is only read and \emph{y} is only written within Task 1.
%Now, suppose three tasks have been virtualized into a single
%one at run-time, namely Task 4: since \emph{x} is now first
%\emph{read} and then \emph{written}, a WAR dependency on
%\emph{x} is introduced dynamically at
%run-time---unfortunately Task 4 is \emph{no more atomic}
%since its re-execution will not always produce the same
%results.
%
Although WAR dependencies introduced by the \emph{static} task division can be eliminated at compile time~\citep{alpaca}, those introduced \emph{dynamically} at
run-time cannot. To maintain memory consistency in merged tasks, we propose a
mechanism for \emph{memory virtualization} that buffers the updates to non-volatile variables in volatile memory before committing them at the dynamic task boundary.
%
%Therefore, we require a \emph{new execution
%model} that keeps each virtual task atomic by committing the
%modified persistent variables at the boundary defined
%dynamically at run-time---keeping the non-volatile memory
%unmodified upon a power-interrupt and preserving its
%consistency.

@@

%Existing programming and execution models may require more accesses to non-volatile memory (for multi-versioning)~\citep{dino,chain} or may preclude the use of volatile memory~\citep{ratchet}. Effective use of both volatile and non-volatile memory is important for \emph{efficiency}, because non-volatile memory has higher access energy and latency than SRAM~\citep{nvp}, and \emph{generality}, because devices often have much more FRAM than SRAM (e.g., 64 times more in~\citep{wolverine}). How to connect this trade-off with task virtualization remains an open question.

@@

We evaluated \sys end-to-end on a real energy-harvesting device~\citep{wisp} running existing benchmarks ~\citep{chain} and new programs. A comparison to a state-of-the-art task-based system~\citep{chain} showed that \sys recovers
performance on conservative task decompositions missed by systems with static decomposition.
%
%and can flexibly target a variety of platforms without rewriting code.
%
%\todo{Provide concrete numeric results supporting this claim}{Amjad/Przemek}

%In this work we also address the problem of task construction burden imposed on the programmer. That is, we introduce \sys's compiler that transforms arbitrary C code into a graph of tasks that reflects the original program's control- and data-flow constraints. The compiler identifies data shared by multiple of the newly created tasks and instruments reads and writes of those data so that \sys's memory virtualization mechanism keeps those data consistent. While \sys's complier is still \emph{limited in scope}, e.g. is unable to handle instruction jumps, normalized loops and recursions, it already eliminates the labour-intensive process of writing tasks~\citep{chain,alpaca}, i.e., hours of programmer time.
